---
title: "Benchmark"
author: "Klaus Schliep"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
bibliography: phyloregion.bib
vignette: >
   %\VignetteIndexEntry{phylorregion-intro}
   %\VignetteEngine{knitr::rmarkdown}
   %\usepackage[utf8]{inputenc}   
---

## A small benchmark study

In this vignette we will benchmark several common functions for several packages
computing phylogenetic diversity (Faith 1992), beta diversity and phylogenetic 
beta diversity. 

Lets load first a all the packages we are going to use. 
```{r, echo=FALSE}
library(ape)
library(Matrix)
library(bench)
suppressPackageStartupMessages( library(ggplot2) )
# packages we benchmark
suppressPackageStartupMessages( library(phyloregion) )
suppressPackageStartupMessages( library(betapart) )
suppressPackageStartupMessages( library(picante) )
suppressPackageStartupMessages( library(vegan) )
suppressPackageStartupMessages( library(hilldiv) )
suppressPackageStartupMessages( library(BAT) )
```
```{r}
library(ape)
library(Matrix)
library(bench)
library(ggplot2)
# packages we benchmark
library(phyloregion)
library(betapart)
library(picante)
library(vegan)
library(hilldiv)
library(BAT)
```
We will use a small data set which comes with `phyloregion`. 
```{r}
data(africa)
# subset matrix
X_sparse <- africa$comm[1:30, ]
X_sparse <- X_sparse[, colSums(X_sparse)>0]
X_dense <- as.matrix(X_sparse)
Xt_dense <- t(X_dense)

object.size(X_sparse)
object.size(X_dense)
dim(X_sparse)
```
To compare results and start the analysis often we need to reformat the input or
results. E.g. for the `hilldiv` package the community matrix needs to be 
transposed. These transformations influence the timings usually only marginal. 
For the benchmarking we use the package bench as it returns timings and provides
some information about memory allocations.
 

### Beta diversity

For beta diversity there is only a small advantage: 
```{r, beta_diversity, cache=TRUE}
chk_fun <- function(target, current)
    all.equal(target, current, check.attributes = FALSE)

fun_phyloregion <- function(x) as.matrix(phyloregion::beta_diss(x)[[3]])
fun_betapart <- function(x) as.matrix(betapart::beta.pair(x)[[3]])
fun_vegan  <- function(x) as.matrix(vegan::vegdist(x, binary=TRUE)) 
fun_BAT <- function(x) as.matrix(BAT::beta(x, func = "Soerensen")[[1]])
results <- bench::mark(fun_phyloregion(X_sparse), fun_betapart(X_dense), 
                       fun_vegan(X_dense), fun_BAT(X_dense), check=chk_fun)
summary(results)
autoplot(results)
``` 
 
### Phylogenetic diversity (Faith 1992)
The is a large difference in memory allocations and running time for Faith's 
phylogenetic diversity. 
```{r, phylo_diversity, cache=TRUE}
tree <- africa$phylo
tree <- keep.tip(tree, colnames(X_sparse))  

pd_picante <- function(x, tree){
    res <- picante::pd(x, tree)[,1]
    names(res) <- row.names(x)
    res    
}    
pd_hilldiv <- function(x, tree) hilldiv::index_div(x, tree, index="faith")
pd_phyloregion <- function(x, tree) phyloregion::PD(x, tree) 

results <- bench::mark(pd_picante(X_dense, tree),
          pd_hilldiv(Xt_dense,tree=tree),
          pd_phyloregion(X_sparse, tree))
summary(results)
autoplot(results)           
```                       

### Phylogenetic beta diversity
Again there is a large difference in memory allocations and running time for the 
phylogenetic beta diversity. 
```{r, phylobeta, cache=TRUE}
fun_phyloregion <- function(x, tree) phyloregion::phylobeta(x, tree)[[3]]
fun_betapart <- function(x, tree) betapart::phylo.beta.pair(x, tree)[[3]]
fun_picante <- function(x, tree) 1 - picante::phylosor(x, tree)
fun_BAT <- function(x, tree) BAT::beta(x, tree, func = "Soerensen")[[1]] 

chk_fun <- function(target, current)
    all.equal(target, current, check.attributes = FALSE)

results <- bench::mark(fun_picante(X_dense, tree), 
                       fun_betapart(X_dense, tree),
                       fun_BAT(X_dense, tree),
                       fun_phyloregion(X_sparse, tree), check=chk_fun)
summary(results)
autoplot(results)
```
Note that the picante function returns a similarity matrix while the function in 
betapart and phyloregion return a dissimilarity matrix. 

## Session Infomation
```{r, eval=TRUE}
sessionInfo()
```

