---
title: "An introduction to the phyloregion package"
author: "Barnabas H. Daru, Piyal Karunarathne & Klaus Schliep"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
bibliography: phyloregion.bib
csl: global-ecology-and-biogeography.csl
vignette: >
   %\VignetteIndexEntry{phyloregion-intro}
   %\VignetteEngine{knitr::rmarkdown}
   %\usepackage[utf8]{inputenc}   
---
---
nocite: | 
   @Daru2020, @Daru2017
---

## 1. Installation

`phyloregion` is a free R package hosted on [GitHub](https://github.com/darunabas/phyloregion). To install `phyloregion`,
with the help from the `remotes` package, type the following commands in R: 
```
if (!requireNamespace("remotes", quietly = TRUE)) 
    install.packages("remotes") 
remotes::install_github("darunabas/phyloregion")
```
When installed, load the package in R:
```{r}
library(phyloregion)
```

## 2. Overview and general workflow of `phyloregion`
The workflow of the `phyloregion` package demonstrates steps from preparation 
of different types of data to visualizing the results of biogeographical
regionalization, together with tips on selecting the optimal method for 
achieving the best output, depending on the types of data used and research
questions.

![Simplified workflow for analysis of biogeographical regionalization using phyloregion. Distribution data is converted to a sparse community matrix. When paired with phylogenetic data, phylobuilder creates a subtree with largest overlap from a species list, thereby ensuring complete representation of missing data; phylocommunity matrix to visualization of results.](workflow.png)

## 3. Input data

### Phylogenies
In R, phylogenetic relationships among species / taxa are often represented as a phylo 
object implemented in the `ape` package [@Paradis2018]. Phylogenies (often in the Newick 
or Nexus formats) can be imported into R with the `read.tree` or `read.nexus` 
functions of the `ape` package.


### Community data

Community data are commonly stored in a matrix with the sites as 
rows and species / operational taxonomic units (OTUs) as columns. The 
elements of the matrix are numeric values indicating the abundance/observations 
or presence/absence (0/1) of OTUs in different sites. In practice, such a matrix can contain many zero values because species are known to generally have unimodal distributions along environmental gradients [@TerBraak2004], and storing and analyzing every single element of that matrix can be computationally challenging and expensive. 

`phyloregion` differs from other R packages (e.g. vegan [@vegan], picante [@Kembel2010]
or betapart [@Baselga2012]) in that the data are not stored in a (dense) `matrix` or `data.frame` but as a sparse matrix making use of the infrastructure provided by the Matrix package [@Matrix]. A sparse matrix is a matrix with a high proportion of zero entries [@Duff1977], of which only the non-zero entries are stored and used for downstream analysis.

A sparse matrix representation has two advantages. First the community matrix 
can be stored in a much memory efficient manner, allowing analysis of larger 
datasets. Second, for very large datasets spanning thousands of taxa and spatial scales, 
computations with a sparse matrix are often much faster.  
The `phyloregion` package contains functions to conveniently change between data 
formats. 
```{r, eval=TRUE}
library(Matrix) 
data(africa)
sparse_comm <- africa$comm
dense_comm <- as.matrix(sparse_comm) 
object.size(dense_comm)
object.size(sparse_comm)
```
Here, the data set in the dense matrix representation consumes roughly five times 
more memory than the sparse representation. 

## 4. Analysis

### Phylogenetic beta diversity

`phyloregion` offers a fast means of computing phylogenetic beta diversity, the turnover 
of branch lengths among sites, making use of and improving on the infrastructure provided by the `betapart` package [@Baselga2012].  
```{r}
tree <- africa$phylo
tree <- ape::keep.tip(tree, colnames(sparse_comm))
pb <- phylobeta(sparse_comm, tree)
```

## Session Information
```{r, eval=TRUE}
sessionInfo()
```

## REFERENCES

