---
title: "An introduction to the phyloregion package"
author: "Klaus Schliep"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
bibliography: phyloregion.bib
nocite: '@Daru2017'
vignette: >
   %\VignetteIndexEntry{phyloregion-intro}
   %\VignetteEngine{knitr::knitr}
   %\usepackage[utf8]{inputenc}   
---


## Installing phyloregion


phyloregion is an free R package hosted on GitHub. To install the phyloregion 
with the help from the remotes package type the following commands in R: 
```
if (!requireNamespace("remotes", quietly = TRUE)) 
    install.packages("remotes") 
devtools::install_github("darunabas/phyloregion")
```
Once the package is installed you load the package inside R:
```{r}
library(phyloregion)
```

## Data formats

### Phylogenies
Phylogenetic relationships among species / taxa are represented in the phylo 
format implemented in the ape package [@Paradis2018]. Phylogenies in Newick or 
Nexus format it can be imported into R with the `read.tree` or `read.nexus` 
functions of the ape package.



### Community data

Community data are commonly stored in a matrix with the sites / samples in the 
rows and the species / otu's (operational taxonomic units) in the columns. The 
elements of the matrix are numeric values indicating the abundance/observations 
or resence/absence (0/1) of otu's in different samples. Very often we observe 
that the majority of the the entries in such a comunity matrix is zero. 
Phyloregion differs from other R packages like vegan [@vegan], picante 
[@Kembel2010] or betapart [@Baselga2012] that the data are not stored in a 
(dense) `matrix` or `data.frame` but as a sparse matrix making use 
infrastructure provided by the Matrix package [@Matrix]. 

Th advantages are that usually that the community matrix can be stored much more 
memory efficient and also computatuions with a sparse matrix are also 
considerably faster. The phyloregion and Martrix package contains some functions 
to conveniently change between formats. 
```{r, eval=TRUE}
library(Matrix) 
data(africa)
comm_sparse <- africa$comm
comm_dense <- as.matrix(comm_sparse) 
object.size(comm_dense)
object.size(comm_sparse)
```
The data set the dense matrix representation consumes roughly five times more 
memory than the sparse representation. 

## Analysis

### Phylogenetic beta diversity

Phyloregion offers a fast way to compute the phylogenetic beta diversity, making 
use and improving on the infrastructure provided by the betapart package 
[@Baselga2012].  
```{r}
tree <- africa$phylo
tree <- ape::keep.tip(tree, colnames(comm_sparse))
pb <- phylobeta(comm_sparse, tree)
```


## Session Infomation
```{r, eval=TRUE}
sessionInfo()
```

## References



