---
title: "An introduction to the phyloregion package"
author: "Klaus Schliep"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
bibliography: phyloregion.bib
nocite: '@Daru2017'
vignette: >
   %\VignetteIndexEntry{phyloregion-intro}
   %\VignetteEngine{knitr::knitr}
   %\usepackage[utf8]{inputenc}   
---


## Installing phyloregion


`phyloregion` is an free R package hosted on GitHub. To install the phyloregion 
with the help from the remotes package type the following commands in R: 
```
if (!requireNamespace("remotes", quietly = TRUE)) 
    install.packages("remotes") 
devtools::install_github("darunabas/phyloregion")
```
Once the package is installed you load the package inside R:
```{r}
library(phyloregion)
```

## Data formats

### Phylogenies
Phylogenetic relationships among species / taxa are represented in the phylo 
format implemented in the ape package [@Paradis2018]. Phylogenies often in the Newick 
or Nexus format can be imported into R with the `read.tree` or `read.nexus` 
functions of the `ape` package.


### Community data

Community data are commonly stored in a matrix with the sites in the 
rows and the species / operational taxonomic units (OTUs) in the columns. The 
elements of the matrix are numeric values indicating the abundance/observations 
or presence/absence (0/1) of OTUs in different sites. In practice, such a matrix can contain many zero values because species are known to generally have unimodal distributions along environmental gradients (ter Braak & Prentice, 1988), and storing and analyzing every single element of that matrix can be computationally challenging and expensive. 

`phyloregion` differs from other R packages like vegan [@vegan], picante 
[@Kembel2010] or betapart [@Baselga2012] in that the data are not stored in a 
(dense) `matrix` or `data.frame` but as a sparse matrix making use of the 
infrastructure provided by the Matrix package [@Matrix]. A sparse matrix is a matrix with a high proportion of zero entries (Duff 1977), of which only the non-zero entries
are stored and utlized for downstream analysis.

The are two advantages of the sparse representation: First the community matrix 
can be stored much more memory efficient, which allows to analyze larger 
datasets and second computations with a sparse matrix are often much faster. 
The `phyloregion` package contains functions to conveniently change between data 
formats. 
```{r, eval=TRUE}
library(Matrix) 
data(africa)
comm_sparse <- africa$comm
comm_dense <- as.matrix(comm_sparse) 
object.size(comm_dense)
object.size(comm_sparse)
```
Here, the data set in the dense matrix representation consumes roughly five times 
more memory than the sparse representation. 

## Analysis

### Phylogenetic beta diversity

`phyloregion` offers a fast way to compute phylogenetic beta diversity, the turnover 
of branch lengths among sites, making use and improving on the infrastructure provided by the betapart package [@Baselga2012].  
```{r}
tree <- africa$phylo
tree <- ape::keep.tip(tree, colnames(comm_sparse))
pb <- phylobeta(comm_sparse, tree)
```

## Session Infomation
```{r, eval=TRUE}
sessionInfo()
```

## References



